---
title: "MasterControl EDA - Group 2"
output: 
  html_document:
    number_sections: yes
    toc: yes
editor_options: 
  chunk_output_type: inline
---

# Introduction

MasterControl's Manufacturing Solutions product (Mx) is significantly underperforming its Quality Solutions product (Qx). The company believes it may not be targeting the right companies or the right people within those companies, leading to missed revenue opportunities and wasted marketing expenditures.

Conversion predictions will be generated using a classification model. The model will use historical sale leads and data collected from MasterControl. Then we will analyze patterns produced by the models to identify similarities between higher-converting accounts. 

The purpose of this EDA notebook is to explore the relationships between different variables in the data. We will explore questions like what are the common characteristics of contacts/companies that are best prospects for Mx and how they differ from Qx products.

# Questions

Questions about that data that is being explored include:
- What data types are in the data?
- Are there missing values?
- What are the different industry types? 
- Do any of those industries have higher conversion rates?
- Are there certain features that are more common in clients who convert?
- Is there overlap between industries that convert for Mx v.s. Qx?

# Description of the data

```{r explore the data}

# Import packages
suppressMessages(suppressWarnings({
library(tidyverse)
library(dplyr)
library(stringr)
library(ggplot2)
  }))

# Import csv file
master <- read.csv("QAL Performance for MSBA.csv")

# Quick look at the data
master |>
  str()

master |>
  summary()

# Checking the proportion of each target variable value
prop.table(table(master$next_stage__c))

```

There are 16,815 observations with 14 variables that are all characters. If we add all of the observations that have progressed (SQL, SQO, and Won) then that makes up about 17.7% of the data. 

Right now, qal.cohort.date is stored as characters, we will have to convert this to a date format.

```{r convert qualification date format}

master$qal.cohort.date <- ymd(master$qal.cohort.date)
  
```

# Discussion of missing data

```{r filtering to missing data}

# Checking for blanks in each column
NA_col <- colSums(master == "", na.rm = TRUE)

# Check total blanks across all data
NA_total <- sum(master == "", na.rm = TRUE)

NA_col
NA_total

```

There are 8 variables that have blank values, and a total of 20,868 cells that were left blank.

### Converting blanks to NA

We have identified the blanks so now we can convert them to NA.

```{r convert blanks to NA}

master_clean <- master |>
  mutate(across(where(is.character), ~ na_if(., "")))

# Check for blanks again
colSums(master_clean == "", na.rm = TRUE)

# Calculating the percent of NAs in each column
master_clean |>
  summarise(across(everything(),
                   ~ round(mean(is.na(.)) * 100, 2)))

```

acct.target.industry, contact.lead.id, next_stage__c, acc.territory.rollup, and acct.tier.rollup all have less than 3% of NAs in its respective columns. Since this makes up such a miniscule amount, we will delete these rows. 

```{r deleting rows with very few NAs}

master_clean <- master_clean |>
  filter(!is.na(acct.target.industry),
         !is.na(contact.lead.id),
         !is.na(next_stage__c),
         !is.na(acct.territory.rollup),
         !is.na(acct.tier.rollup))

# Checking to see if NAs were dropped in those columns
master_clean |>
  summarise(across(everything(),
                   ~ round(mean(is.na(.)) * 100, 2)))

```

acct.primary.site.function, acct.manufacturing.model, and contact.lead.title have more than 35% of its data NAs. Since this is such a large amount, we will keep these rows because they could still hold valuable information. There are no numeric values in the dataset so we can convert NA to "Missing" instead for categorical analysis.

```{r converting NA to Missing}

master_clean <- master_clean |>
  mutate(across(where(~ is.character(.)), 
                ~ ifelse(is.na(.), "Missing", as.character(.))))

# Checking NAs
master_clean |>
  summarise(across(everything(),
                   ~ round(mean(is.na(.)) * 100, 2)))

```
Now there are no NAs in the dataset. They have been replaced by "Missing" instead and we will treat them as its own category.

### Grouping unique values to larger categories

Another important thing about this dataset is that there are many unique values for each column. This creates high cardinality and can make it hard for us to model our data. We will group these categories into larger groups to make them more meaningful.

```{r grouping categories}

# Grouping SQL, SQO, and Won as Success and the rest as Failure
master_clean <- master_clean |>
  mutate(success_binary = ifelse(
    next_stage__c %in% c("SQL", "SQO", "Won"),
    1,
    0
  ))

# Checking class balance
prop.table(table(master_clean$success_binary))

# Converting all contact lead titles to uppercase for standardization
master_clean <- master_clean |>
  mutate(contact.lead.title = str_trim(contact.lead.title),
         contact.lead.title = str_to_upper(contact.lead.title))

# Grouping contact lead titles to more general categories
master_clean <- master_clean |>
  mutate(title_group = case_when(
    str_detect(contact.lead.title, "CHIEF|CEO|CFO|CTO|COO|CIO|PRESIDENT|OWNER|FOUNDER|EXECUTIVE|DIR") ~ "Executive",
    str_detect(contact.lead.title, "VP|VICE PRESIDENT|HEAD") ~ "VP / Head",
    str_detect(contact.lead.title, "DIRECTOR") ~ "Director",
    str_detect(contact.lead.title, "MANAGER|SUPERVISOR|MANAGE|MANGER") ~ "Manager",
    str_detect(contact.lead.title, "ENGINEER|SCIENTIST|TECHNICAL|DEVELOPER|LEAD|TECHNICIAN|ARCHITECT") ~ "Technical",
    str_detect(contact.lead.title, "ANALYST") ~ "Analyst",
    str_detect(contact.lead.title, "SPECIALIST|ASSOCIATE|COORDINATOR") ~ "Specialist",
    str_detect(contact.lead.title, "CONSULTANT|CONSULTING") ~ "Consultant",
    str_detect(contact.lead.title, "MISSING") ~ "Unknown",
    str_detect(contact.lead.title, "PROFESSOR|RESEARCHER|RESEARCH|LECTURER|TEACHER|PROF") ~ "Academia",
    str_detect(contact.lead.title, "SALES") ~ "Sales",
    TRUE ~ "Other"
  ))

# Checking title group distribution
prop.table(table(master_clean$title_group))

# Converting all primary site functions to uppercase for standardization
master_clean <- master_clean |>
  mutate(acct.primary.site.function = str_trim(acct.primary.site.function),
         acct.primary.site.function = str_to_upper(acct.primary.site.function))

# Grouping acct primary site function
master_clean <- master_clean |>
  mutate(site_function = case_when(
    str_detect(acct.primary.site.function, "OSD|STERILE INJECTABLES|API|BIOLOGICS|CELL THERAPIES") ~ "Active Pharma Manufacturing",
    str_detect(acct.primary.site.function, "IMPLANTABLE DEVICES|IVD|INSTRUMENTS|PHARMA|HEALTH|MEDICAL|SURGICAL|PATIENT|DRUG|CARE|DIAGNOSTIC|BLOOD|VACCINE") ~ "Medical Device Manufacturing",
    str_detect(acct.primary.site.function, "SERVICES|CONSULTING|NON-MANUFACTURING") ~ "Non-Manufacturing",
    str_detect(acct.primary.site.function, "FOOD|BEVERAGE|NUTRACEUTICAL|COSMETICS|CHEMICALS") ~ "Other Manufacturing",
    str_detect(acct.primary.site.function, "NOT ENOUGH INFO FOUND|MISSING") ~ "Unknown",
    TRUE ~ "Other"
  ))

prop.table(table(master_clean$site_function))

```

# Exploratory visualizations and /or summary tables

```{r target variable class balance}

table(master_clean$success_binary)
prop.table(table(master_clean$success_binary))

# Distinction between Mx and Qx successes
master_clean |>
  group_by(solution_rollup) |>
  summarise(
    total = n(),
    successes = sum(success_binary),
    success_rate = mean(success_binary)
  )

```

There are 2,994 recorded successes and 13,209 failures. Earlier we calculated that 17.7% of the dataset were successes but after cleaning the data, that has increased to 18.5%.

Within Mx solutions alone, we have a success rate of 12.8% while Qx has 20.3%, which is expected.

```{r distribution of title groups by success}

master_clean |>
ggplot(aes(x = title_group, fill = factor(success_binary))) +
  geom_bar(position = "dodge") +
  coord_flip() +
  facet_wrap(~ solution_rollup) +
  labs(fill = "Success",
       x = "Title Group",
       y = "Count",
       title = "Distribution of Title Groups by Success")

```

A lot of the contact lead titles were left blank so there are many Unknowns that make up a lot of the data. If we ignore that category though, we see that many of the successes are from Manager/Executive roles. The next best category are Technical roles.

```{r distribution of site function by success}

master_clean |>
ggplot(aes(x = site_function, fill = factor(success_binary))) +
  geom_bar(position = "dodge") +
  geom_text(
    stat = "count",
    aes(label = ..count..),
    position = position_dodge(width = 0.9),
    hjust = -0.2
  ) +
  coord_flip() +
  facet_wrap(~ solution_rollup) +
  labs(fill = "Success",
       x = "Site Function",
       y = "Count",
       title = "Distribution of Site Functions by Success")

```

Much like the lead titles, there are also a lot of data that is missing and make up the majority of failures and successes in the dataset. If we ignore the Unknown category, the next biggest contributors are medical device manufacturing and active pharma site functions.

```{r distribution of target industry by success}

master_clean |>
ggplot(aes(x = acct.target.industry, fill = factor(success_binary))) +
  geom_bar(position = "dodge") +
  coord_flip() +
  facet_wrap(~ solution_rollup) +
  labs(fill = "Success",
       x = "Target Industry",
       y = "Count",
       title = "Distribution of Target Industries by Success")

```

Most successes are seen within the pharma and biotech, non-life science, and medical device industries. 

```{r success rate by month}

# Creating a month variable
master_clean <- master_clean |>
  mutate(month = floor_date(qal.cohort.date, "month"))

# Success rate by month filtered to only Mx products
master_clean |>
  filter(solution_rollup == "Mx") |>
  group_by(month) |>
  summarise(
    total = n(),
    success_rate = mean(success_binary)
  ) |>
  ggplot(aes(x = month, y = success_rate)) +
  geom_line() +
  geom_point() +
  labs(title = "Mx Success Rate by Month",
       x = "Month",
       y = "Success Rate") +
  theme_minimal()


```

Mx success rate seem to hover between 15% - 25% between 2024 and 2025. It seems to have performed better in 2024 than it did in 2025, but not by a lot. 

```{r success rate by territory}

master_clean |> 
  filter(solution_rollup == "Mx") |>
  ggplot(aes(x = acct.territory.rollup, fill = factor(success_binary))) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Mx Success Rate by Territory",
       y = "Proportion",
       fill = "Success") +
  theme_minimal()


```

Mx products predominately show successes in Japan and the Americas, with lower presence in EMEA and APAC/Oceania markets. 

Gaurav Mishra

# Conversion by Marketing Channel
# Best performing channels --> highest quality leads 
- Direct/Inbound
- SEP
- Online Ads

```{r}
df_clean %>%
  group_by(last_tactic_campaign_channel) %>%
  summarise(
    total_leads = n(),
    sqls = sum(next_stage_c == "SQL"),
    conversion_rate = sqls / total_leads
  ) %>%
  arrange(desc(conversion_rate))
```

# Conversion by industry

# Highest SQL Conversion Rate
- Blood & biologics
- Medical devices
- Pharma & BioTech 

```{r}
df_clean %>%
  group_by(acct_target_industry) %>%
  summarise(
    total_leads = n(),
    sqls = sum(next_stage_c == "SQL"),
    conversion_rate = sqls / total_leads
  ) %>%
  arrange(desc(conversion_rate))

```

# Conversion by Account Tier 
- Small accounts convert best, medium, then large

```{r}
df_clean %>%
  group_by(acct_tier_rollup) %>%
  summarise(
    total_leads = n(),
    sqls = sum(next_stage_c == "SQL"),
    conversion_rate = sqls / total_leads
  ) %>%
  arrange(desc(conversion_rate))

```

# Industry x Tier
```{r}
df_clean %>%
  group_by(acct_target_industry, acct_tier_rollup) %>%
  summarise(
    total_leads = n(),
    sqls = sum(next_stage_c == "SQL"),
    conversion_rate = sqls / total_leads,
    .groups = "drop"
  ) %>%
  arrange(desc(conversion_rate))

```
# Heatmap
```{r}
library(scales)

heat_data <- df_clean %>%
  group_by(acct_target_industry, acct_tier_rollup) %>%
  summarise(
    conversion_rate = mean(next_stage_c == "SQL"),
    .groups = "drop"
  )

ggplot(heat_data, 
       aes(acct_tier_rollup, acct_target_industry, fill = conversion_rate)) +
  geom_tile() +
  scale_fill_gradient(
    low = "white",
    high = "steelblue",
    labels = scales::percent_format()
  ) +
  labs(
    title = "Conversion Rate Heatmap",
    x = "Account Tier",
    y = "Industry",
    fill = "Conversion"
  ) +
  theme_minimal()
```

# Scatter Plot of Qx and Mx Conversion
```{r Qx v.s. Mx Conversion by Industry}
# Conversion
conv_compare <- QAL_Perform_clean |>
  filter(solution_rollup %in% c("Mx", "Qx")) |>
  group_by(solution_rollup, `acct target industry`) |>
  summarise(
    conv_rate = mean(converted, na.rm = TRUE),
    n_leads = n(),
    .groups = "drop"
  ) |>
  select(solution_rollup, `acct target industry`, conv_rate) |>
  pivot_wider(
    names_from = solution_rollup,
    values_from = conv_rate)

# Scatterplot
ggplot(conv_compare, aes(x = Qx, y = Mx)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_text(aes(label = `acct target industry`),
            hjust = 0, nudge_x = 0.005, size = 3) +
  labs(
    title = "Industry Conversion: Qx vs Mx",
    x = "Qx Conversion Rate",
    y = "Mx Conversion Rate")
```
# Analysis Venn Diagram Mx and Qx

```{r }
# Industries with Mx conversions
mx_industries <- QAL_Perform_clean |>
  filter(solution_rollup == "Mx", converted == TRUE) |>
  pull(`acct target industry`) |>
  unique()

# Industries with Qx conversions
qx_industries <- QAL_Perform_clean |>
  filter(solution_rollup == "Qx", converted == TRUE) |>
  pull(`acct target industry`) |>
  unique()

# Industry List
industry_list <- list(
  Mx = mx_industries,
  Qx = qx_industries)

ggVennDiagram(industry_list) +
  labs(title = "Overlap of Industries with Conversions: Mx vs Qx")

# Overlap Values
intersect(mx_industries, qx_industries)
```

# Results

Our findings show that Mx product successes are most often achieved when the contact lead titles are at the executive or manager level within medical device manufacturing and active pharma site functions. The target industries for these customers are pharma and biotech, non-life science, and medical devices. Mx's success rates are pretty stagnant in 2024 and 2025, with slightly lower performance in 2025. Lastly, the Mx product is most successful in Japan, the Americas, and EMEA territories.

Segment-level interaction analysis shows that small-tier companies outperform medium and large accounts across several industries, particularly within core life sciences segments. This pattern implies that prioritizing smaller and mid-market firms may drive higher pipeline efficiency than focusing heavily on enterprise accounts.

Blood & Biologics is the highest-converting industry for Mx, followed by Medical Device. The industry ranking is the same for both Mx and Qx, suggesting that industry alone does not explain the performance gap. For Mx leads, CMOs show the highest conversion rate followed by CDMO and In-House. CROs and Research Institutes underperform substantially, which aligns with expectations, as companies that do not manufacture products are weaker Mx prospects. For Qx, the rankings look different: non-manufacturing models like Consumer Packaged Goods, Distribution/Wholesale, and Software/Technology Development convert well. While these same models show few or no Mx leads. This confirms that there is a fundamental difference in the buyer profile between the two products.

Although our EDA highlights key trends within the data, there is still a substantial portion where there are missing/unknown values. This limits the insights we can draw from this and we will have to take this in consideration as we move forward into the modeling part of our project. 
